Data organization in spreadsheets

### 2017-02-24 manuscript.Rmd
Revisions in response to reviewer 1

Use consistent codes for categorical variables. For a
categorical variable like *sex*, use a single common value for
males (e.g. "`male`") and a single common value for females
(e.g. "`female`"). Don't sometimes write "`M`", sometimes "`male`",
and sometimes "`Male`". Pick one and stick to it.

Use a single fixed code for any missing values. We prefer to have
every cell filled in,
so that one can distinguish between
truly missing values and unintentionally missing
values. R users prefer "`NA`". You
could also use a hyphen. But stick with a single value
throughout your data. Definitely don't use a numeric value like
`-999` or `999`; it is easy to miss that it is intended to be
missing. Also, don't insert a note in place of the data,
explaining why it is missing. Rather, make a separate column with
such notes.

Use consistent variable names. If in one file (say the first
batch of subjects), you have a variable called "`Glucose_10wk`",
then call it exactly that in other files (say for other batches of
subjects). If it is variably called "`Glucose_10wk`",
"`gluc_10weeks`", and "`10 week glucose`", then downstream the
data analyst will have to work out that these are all really the
same thing.

Use consistent subject identifiers. If sometimes it is "`153`" and
sometimes "`mouse153`" and sometimes "`mouse-153F`" and sometimes
"`Mouse153`", there is going to be extra work to figure out who is
who.

Use a common data layout in multiple files. If your data are in multiple files
and you use different layouts in different files, it will be extra work for the
analyst to combine the files into one dataset for analysis. With a consistent
structure, it will be easy to automate this process.

Use consistent file names. Have some system for naming files. If one file is
called "`Serum_batch1_2015-01-30.csv`", then don't call the file for the next
batch "`batch2_serum_52915.csv`" but rather use "`Serum_batch2_2015-05-29.csv`".
Keeping a consistent file naming scheme will help ensure that your files remain
well organized, and it will make it easier to batch process the files if you
need to.

Use a single common format for all dates, preferably with the standard
format `YYYY-MM-DD`, for example `2015-08-01`. If sometimes you write `8/1/2015`
and sometimes `8-1-15`, you are asking for trouble.

Use consistent phrases in your notes. If you have a separate
column of notes (for example, "`dead`" or "`lo off curve`"), be
consistent in what you write. Don't sometimes write "`dead`" and
sometimes "`Dead`", or sometimes "`lo off curve`" and sometimes
"`off curve lo`".

Be careful about extra spaces within cells. A blank cell is
different than a cell that contains a single space. And "`male`"
is different from "` male `" (that is, with spaces at the
beginning and end). These can be a headache later on.

@@

@@

```{r 01_bad_dates, fig.cap="A spreadsheet with inconsistent date formats."}
mat <- data.frame(Date=c("", "", "12/6/2005", rep("", 4)),
                  "Assay date"=c("12/9/05", "12/9/05", "e", "e", "e", "1/11/2006", "1/11/2006"),
                  "Weight"=c(54.9, 45.3, 47, 45.7, 52.9, 46.1, 38.6),
                  stringsAsFactors=FALSE, check.names=FALSE)
excel_fig(mat, cellwidth=90)
```

@@

```{r 02_blank_cells, fig.width=12, fig.cap="Examples of spreadsheets with blank cells. **A:** A spreadsheet where only the first of several repeated values was included. **B:** A spreadsheet with a complicated layout and some implicit column headers."}
layout(cbind(1,2), width=c(8,10))
mat <- data.frame(id=101:107,
                  date=c("2015-06-14", "", "2015-06-18", "", "", "2015-06-20", ""),
                  glucose=myround(c(149.3, 95.3, 97.5, 117.0, 108.0, 149.0, 169.4), 1),
                  stringsAsFactors=FALSE)
excel_fig(mat, cellwidth=110, mar=c(0.1, 0.1, 1.6, 1.1))
mtext(side=3, adj=0, "A", font=2, cex=1.2, line=0.5)

@@

The best layout for your data within in a spreadsheet is as a big
rectangle with rows corresponding to subjects and columns
corresponding to variables. The first row should contain variable
names. (*Please don't use more than one row for the variable names.*)
An example of a rectangular layout is shown in Figure 4.

```{r 04_example_rectangle, fig.height=3, fig.cap="An example spreadsheet with a rectangular layout."}
set.seed(20150803)
mat <- data.frame(id=101:105,
                  sex=sample(c("Male", "Female"), 5, replace=TRUE),
                  glucose=myround(runif(5, 70, 150), 1),
                  insulin=myround(runif(5, 0.5, 1.5), 2),
                  triglyc=myround(runif(5, 70, 300), 1),
                  stringsAsFactors=FALSE)

@@

Some data sets won't fit nicely into a rectangle, but they will usually
fit into a set of rectangles, in which case you can make a set of Excel files,
each with a rectangle of data. Some data don't even fit into a set of
rectangles, but then maybe spreadsheets are not the best format for
them, as spreadsheets are inherently rectangular.

@@

```{r 05_non_rectangular, fig.height=8, fig.width=12, fig.cap="Examples of spreadsheets with non-rectangular layouts."}
layout(rbind(1:2, 3:4), height=c(6,10))

@@

```{r 07_two_header_rows, fig.width=10, fig.cap="A spreadsheet with two header rows."}
mat6 <- rbind(c("", "", "week 4", "", "", "week 6", "", "", "week 8", "", ""),
              c("Mouse ID", "SEX", rep(c("date", "weight", "glucose"), 3)),
              c("3005", "M", "3/30/2007",  19.3, 635.0, "4/11/2007",  31,   460.7, "4/27/2007",  39.6, 530.2),
              c("3017", "M", "10/6/2006",  25.9, 202.4, "10/19/2006", 45.1, 384.7, "11/3/2006",  57.2, 458.7),
              c("3434", "F", "11/22/2006", 26.6, 238.9, "12/6/2006",  45.9, 378.0, "12/22/2006", 56.2, 409.8),
              c("3449", "M", "1/5/2007",   27.5, 121.0, "1/19/2007",  42.9, 191.3, "2/2/2007",   56.7, 182.5),
              c("3499", "F", "1/5/2007",   19.8, 220.2, "1/19/2007",  36.6, 556.9, "2/2/2007",   43.6, 446.0))

@@

>>>>

_Use consistent codes for categorical variables._ For a
categorical variable like *sex*, use a single common value for
males (e.g. "`male`") and a single common value for females
(e.g. "`female`"). Don't sometimes write "`M`", sometimes "`male`",
and sometimes "`Male`". Pick one and stick to it.

_Use a consistent fixed code for any missing values._ We prefer to have
every cell filled in,
so that one can distinguish between
truly missing values and unintentionally missing
values. R users prefer "`NA`". You
could also use a hyphen. But stick with a single value
throughout your data. Definitely don't use a numeric value like
`-999` or `999`; it is easy to miss that it is intended to be
missing. Also, don't insert a note in place of the data,
explaining why it is missing. Rather, make a separate column with
such notes.

_Use consistent variable names._ If in one file (say the first
batch of subjects), you have a variable called "`Glucose_10wk`",
then call it exactly that in other files (say for other batches of
subjects). If it is variably called "`Glucose_10wk`",
"`gluc_10weeks`", and "`10 week glucose`", then downstream the
data analyst will have to work out that these are all really the
same thing.

_Use consistent subject identifiers._ If sometimes it is "`153`" and
sometimes "`mouse153`" and sometimes "`mouse-153F`" and sometimes
"`Mouse153`", there is going to be extra work to figure out who is
who.

_Use a consistent data layout in multiple files._ If your data are in multiple files
and you use different layouts in different files, it will be extra work for the
analyst to combine the files into one dataset for analysis. With a consistent
structure, it will be easy to automate this process.

_Use consistent file names._ Have some system for naming files. If one file is
called "`Serum_batch1_2015-01-30.csv`", then don't call the file for the next
batch "`batch2_serum_52915.csv`" but rather use "`Serum_batch2_2015-05-29.csv`".
Keeping a consistent file naming scheme will help ensure that your files remain
well organized, and it will make it easier to batch process the files if you
need to.

_Use a consistent format for all dates,_ preferably with the standard
format `YYYY-MM-DD`, for example `2015-08-01`. If sometimes you write `8/1/2015`
and sometimes `8-1-15`, you are asking for trouble.

_Use consistent phrases in your notes._ If you have a separate
column of notes (for example, "`dead`" or "`lo off curve`"), be
consistent in what you write. Don't sometimes write "`dead`" and
sometimes "`Dead`", or sometimes "`lo off curve`" and sometimes
"`off curve lo`".

_Be careful about extra spaces within
cells._ A blank cell is different than a cell that contains a single space. And
"`male`" is different from "` male `" (that is, with spaces at the beginning and
end). These can be a headache later on.

@@

Another way to force Excel to treat dates as text is to begin the date with an
apostrophe, like this: `'2014-06-14`
(see <http://bit.ly/twitter_apos>). Excel will treat the
cells as text, but the apostrophe will not appear when you view the spreadsheet
or export it to other formats. This is a handy trick, but it requires impeccable
diligence and consistency.

@@

```{r 01_bad_dates, fig.cap="A spreadsheet with inconsistent date formats. This spreadsheet does not adhere to our recommendations for consistency of date format."}
mat <- data.frame(Date=c("", "", "12/6/2005", rep("", 4)),
                  "Assay date"=c("12/9/05", "12/9/05", "e", "e", "e", "1/11/2006", "1/11/2006"),
                  "Weight"=c(54.9, 45.3, 47, 45.7, 52.9, 46.1, 38.6),
                  stringsAsFactors=FALSE, check.names=FALSE)
excel_fig(mat, cellwidth=90)
```

@@

```{r 02_blank_cells, fig.width=12, fig.cap="Examples of spreadsheets that violate the 'no empty cells' recommendation. **A:** A spreadsheet where only the first of several repeated values was included. **B:** A spreadsheet with a complicated layout and some implicit column headers. For a tidy version of this data, see Figure 3."}
layout(cbind(1,2), width=c(8,10))
mat <- data.frame(id=101:107,
                  date=c("2015-06-14", "", "2015-06-18", "", "", "2015-06-20", ""),
                  glucose=myround(c(149.3, 95.3, 97.5, 117.0, 108.0, 149.0, 169.4), 1),
                  stringsAsFactors=FALSE)
excel_fig(mat, cellwidth=110, mar=c(0.1, 0.1, 1.6, 1.1))
mtext(side=3, adj=0, "A", font=2, cex=1.2, line=0.5)

@@

The best layout for your data within in a spreadsheet is as a single big
rectangle with rows corresponding to subjects and columns corresponding to
variables. The first row should contain variable names.
(*Please don't use more than one row for the variable names.*)
An example of a rectangular layout is shown in Figure 4.

```{r 04_example_rectangle, fig.height=3, fig.cap="An example spreadsheet with a rectangular layout. This layout will aid future analyses."}
set.seed(20150803)
mat <- data.frame(id=101:105,
                  sex=sample(c("Male", "Female"), 5, replace=TRUE),
                  glucose=myround(runif(5, 70, 150), 1),
                  insulin=myround(runif(5, 0.5, 1.5), 2),
                  triglyc=myround(runif(5, 70, 300), 1),
                  stringsAsFactors=FALSE)

@@

Some data sets won't fit nicely into a single rectangle, but they will usually
fit into a set of rectangles, in which case you can make a set of Excel files,
each with a rectangle of data. It is best to keep each rectangle in its own
file; tables scattered around a worksheet are difficult to work with, and they
make it hard to export the data to CSV files (which we will discuss shortly).

Some data don't even fit into a set of rectangles, but then maybe spreadsheets
are not the best format for them, as spreadsheets are inherently rectangular.

@@

```{r 05_non_rectangular, fig.height=8, fig.width=12, fig.cap="Examples of spreadsheets with non-rectangular layouts. These layouts are likely to cause problems in analysis."}
layout(rbind(1:2, 3:4), height=c(6,10))

@@

```{r 07_two_header_rows, fig.width=10, fig.cap="A spreadsheet with two header rows. It is better to have a single header row (see Figure 8)."}
mat6 <- rbind(c("", "", "week 4", "", "", "week 6", "", "", "week 8", "", ""),
              c("Mouse ID", "SEX", rep(c("date", "weight", "glucose"), 3)),
              c("3005", "M", "3/30/2007",  19.3, 635.0, "4/11/2007",  31,   460.7, "4/27/2007",  39.6, 530.2),
              c("3017", "M", "10/6/2006",  25.9, 202.4, "10/19/2006", 45.1, 384.7, "11/3/2006",  57.2, 458.7),
              c("3434", "F", "11/22/2006", 26.6, 238.9, "12/6/2006",  45.9, 378.0, "12/22/2006", 56.2, 409.8),
              c("3449", "M", "1/5/2007",   27.5, 121.0, "1/19/2007",  42.9, 191.3, "2/2/2007",   56.7, 182.5),
              c("3499", "F", "1/5/2007",   19.8, 220.2, "1/19/2007",  36.6, 556.9, "2/2/2007",   43.6, 446.0))

@@

## Acknowledgments

Lance Waller generously provided comments to improve the manuscript.
